VLI - a library for high precision integer and polynomial arithmetic

### 2013-01-28 doc/paper_vli/vli2.tex
vli paper wip

\section{Large integers}
The library provides a class template \verb|integer<n_bits>| to represent signed integers with a fixed size of $128 \le n \le 512$ bits.
Objects of this class behave like the regular C++ type $int$. We implemented all standard integer operations starting from basic arithmetic operations, comparion operations and bit operations, like bit-shift or bit-wise logical operations.
In addition to these operations we also provide a multiply-add function, an extended multiplcation which doubles the number of bits when two integers of the same size get multiplied and an extended add which increases the size of the integer by one 64bit word.
The data of the integer is stored as a 64 bit unsigned integer array, where we use a standard two-complement represnetation of the number.
Due to the fixed size of the integer, we are able to allocate the memory for the data on the stack.
This way we save expensive heap allocations, which would require a manual memory pool management to get efficient allocations.
The basic arithmetic operations like addition and multiplication use standard schoolbook algorithms, since the integers are too small for Toom\cite{Toom} or FFT based approaches like the Sch\"onhage-Strassen algorithm\cite{Schonhage}.
We implemented the schoolbook algorithms in optimized assembly code to make optimal use of the hardware features that are not accessible from C++, like the carry addition \verb|adcq| or the 64bit multiplication \verb|mulq| which calculates the low and hi 64 bit part of the result on the \verb|x86-64| architecture.
Since writing assembly code can result in very long error-prone code, we generate most parts of the assembly code with the C++ preprocessor using the \verb|BOOST_PP| library.

@@

\section{Polynomials}
The second part of the library is a template class 
\begin{verbatim}
    polynomial<CoeffType, Structure<N>, Var0, Var1, Var2, Var3>
\end{verbatim}
for symbolic polynomials in 1 to 4 variables
\begin{equation}
    p(x,y,z,a) = \sum_{i,j,k,l} c_{ijkl} x^i y^j z^k a^l
\end{equation}
having either a ``dense'' (${i,j,k,l} \le N$) or a ``triangular'' ($i+j+k+l \le N$) structure.
The truncation order $N$ is fixed at compile-time.
As coefficient $c_{ijkl}$ of the polynomial any arithemetic type may be used.

\begin{itemize}
\item {\bf Description:}
\item symbolic polynomial in 1-4 variables.
\item either ``dense'' (${i,j,k,l} \le N$) or ``triangular'' ($i+j+k+l \le N$) structure.
\item any arithmetic type as coefficient $c_{ijkl}$
\item addition, subtraction, multiplication with other polynomials and monomials
\item special multiplication which will double the truncation order
\item {\bf Implementation details:}
\item structure and truncation order $N$ is compile time fixed 
\item function hooks to allow custom optimized kernels for certain coefficient types
\item The symbolic variable names are given as template parameter
\item Automatic matching of variable names of different polynomial types using template meta programming.
\end{itemize}

\section{Optimized inner product}
The hot spot of HTSE are inner products of large vectors with symbolic polynomials as components. The coefficients of the polynomials are large integers.
All polynomials are assumed to have the same truncation order.
The inner product will result in a polynomial of twice the order of the original polynomials
with coefficients of twice the width of the original coefficients.
\paragraph{CPU}
\begin{itemize}
\item We use the single large integer kernels
\item OpenMP parallelization: split vectors into chunks, each thread does part of the inner product. Very regular problem, good load balance.
\item A SIMD implementation to perform 4 large integer operations in parallel on current \verb|x86_64| architectures is not useful, since we had to use 32bit integer operations, which increases the number of multiply instructions for a large integer multiplication by a factor of 4 compared to the 64bit based implementation.
Using current Streaming SIMD Extensions (SSE) allows only two 32bit multiplications yielding 64bit results in a SIMD fashion.
Thus a SIMD implementation using SSE will be two times slower than our serial version using 64 bit integers.
We also neglected the carry bit propagation, which needs to be done manually in SSE, while the sequential version takes advantage of the hardware support for carry bit propagation.
\item The upcoming AVX2 will double the vector width and support four 32bit integer multiplications, but no carry bit support.
\end{itemize}
\paragraph{GPU}
\begin{itemize}
\item Many small tasks of the same form. $\rightarrow$ well suited for GPUs.
\item {\bf Implementation details:}
\item based on 32bit integers due to architecture.
\item Element-wise polynomial multiplication + reduction at the end.
\item Each thread calculates one coefficient $c_{IJKL}$ the product polynomial.
\item Several coefficients $a_{ijkl}$, $b_{i'j'k'l'}$ of the polynomials to be multiplied contribute to the same coefficient $c_{IJKL}$
\begin{equation}
    c_{IJKL} = \sum_{i,j,k,l} a_{ijkl} \cdot b_{I-i,J-j,K-k,L-l}
\end{equation}
\item The number of terms in this sum depends on the orders $I,J,K,L$ of the resulting coefficient $\rightarrow$ load in-balance
%\item {\color{blue} We design an execution plan. We group the coefficients of a fictive result polynomial by packet of 32 } into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}).
\item Since threads within a warp are executed in a lockstep manner, we try to give all threads within a warp the same amount of work. This way the threads have to wait as little as possible for the other threads of the warp to finish.
\item We group the coefficients by 32 into a ``task'' and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}), i.e.\ each thread calculates one coefficient of the resulting polynomial.
\begin{figure}[t]
    \begin{center}
    \mbox{
        \includegraphics[scale=0.37, angle=-90]{coeffs.eps} 
        \includegraphics[scale=0.37, angle=-90]{warp.eps} 
    }
    \caption{Left: Number of contributions to each coefficient of the resulting polynomial (red line) sorted in reverse order and split into tasks. Right: Load balancing of the tasks over the warps.}
    \label{fig:GPU_load_balance}
    \end{center}
\end{figure}
\item Each thread of the warp gets the indices of the resulting coefficient and determines the necessary terms to calculate the output coefficient.
\item Reduction is implemented according to the best practices by NVIDIA (TODO reference)
%\item {\color{blue} This algorithm is efficient whatever the type of polynomial. But the termination of the contributions of an output coefficient can be tricky}
%\item Coalesced memory access?
\item We use the Karatsuba algorithm \cite{Karatsuba1963} for the largest extended multiplication (256 bit to 512 bit) on GPU, because an integer addition requires substantially less cycles than multiplication the GPU.
\item Using NVIDIA parallel thread execution (PTX) assembly language to benefit from hardware carry bit propagation \verb|addc.cc| and multiply-add with carry bit propagation \verb|madc.cc|.
\end{itemize}

@@

\section{Conclusion}
\begin{itemize}
\item TODO
\item License?
\end{itemize}

>>>>

\section{Large integers}
The library provides a template class \verb|integer<n>| to represent signed integers with a fixed size of $128 \le n \le 512$ bits.
Objects of this class behave like the regular C++ type \verb|int|.
We implemented all standard integer operations starting from basic arithmetic operations, comparion operations and bit operations, like bit-shift or bit-wise logical operations.
In addition to these operations we also provide a multiply-add function, an extended multiplcation which doubles the number of bits when two integers of the same size get multiplied and an extended add which increases the size of the integer by one 64 bit word.

The data of the integer is stored as a 64 bit unsigned integer array, where we use a standard two-complement representation of the number.
Due to the fixed size of the integer, we are able to allocate the memory for the data on the stack.
This way we save expensive heap allocations, which would require a manual memory pool management to get efficient allocations.
The basic arithmetic operations like addition and multiplication use standard schoolbook algorithms, since the integers are too small for Toom\cite{Toom} or FFT based approaches like the Sch\"onhage-Strassen algorithm\cite{Schonhage}.
We implemented the schoolbook algorithms in optimized assembly code to make optimal use of the hardware features that are not accessible from C++, like the carry addition \verb|adcq| or the 64 bit multiplication \verb|mulq| which calculates the low and hi 64 bit part of the result on the \verb|x86-64| architecture.
Since writing assembly code can result in very long error-prone code, we generate most parts of the assembly code with the C++ preprocessor using the \verb|BOOST_PP| library.

@@

\section{Polynomials}
The second part of the library is a template class 
\begin{verbatim}
    polynomial<CoeffType, Structure<N>, Var0, Var1, Var2, Var3>
\end{verbatim}
for symbolic polynomials in 1 to 4 variables
\begin{equation}
    p(x,y,z,a) = \sum_{i,j,k,l} c_{ijkl} \cdot x^i y^j z^k a^l
\end{equation}
having either a ``dense'' (${i,j,k,l} \le N$) or a ``triangular'' ($i+j+k+l \le N$) structure.
The truncation order $N$ is fixed at compile-time.
The coefficients $c_{ijkl}$ of the polynomial may have any C++ type/class supporting basic arithmetic operations.
The polynomial itself supports additions, subtractions and multiplications with other polynomials or monomials.
We also provide a special multiplication function returning a polynomial with truncation order $2N$, such that no terms are dropped.
It is also possible to mix and match polynomials having different sets of variables.
The coefficients will be automatically mapped to the corresponding symbols during compile time.
All operations call functions with default implementations for all coefficient types.
These functions may be overloaded by the user to provide optimized implementations for specific coefficient types.
The user may for example add a specialized implementation for \verb|float| using Streaming SIMD Extension intrinsics without touching the library itself.
Except for the general default implementation of the operations the library provides optimized routines for large integer coefficients \verb|integer<n>|.

%\begin{itemize}
%\item {\bf Description:}
%\item symbolic polynomial in 1-4 variables.
%\item either ``dense'' (${i,j,k,l} \le N$) or ``triangular'' ($i+j+k+l \le N$) structure.
%\item any arithmetic type as coefficient $c_{ijkl}$
%\item addition, subtraction, multiplication with other polynomials and monomials
%\item special multiplication which will double the truncation order
%\item {\bf Implementation details:}
%\item structure and truncation order $N$ is compile time fixed 
%\item function hooks to allow custom optimized kernels for certain coefficient types
%\item The symbolic variable names are given as template parameter
%\item Automatic matching of variable names of different polynomial types using template meta programming.
%\end{itemize}

\section{Optimized inner product}
Since the hot spots of our main target application are inner products of large vectors with symbolic polynomials as components, where the coefficients of the polynomials are large integers, we implemented an optimized inner product function for this special case.
All polynomials of the vector are assumed to have the same structure and truncation order $N$ and have the same \verb|integer<n>| type as coefficients.
The inner product will result in a polynomial of twice the order of the original polynomials with coefficients of twice the width \verb|integer<2n>| of the original coefficients.
To employ today's supercomputers as efficiently as possible the inner product is a hybrid CPU/GPU implementation,
where the inner product the vectors are split into a part calculated on the CPU and a part calculated on the GPU.
The ratio of these two parts can be set when building the library.
It is also possible to deactivate either part completely and compile a pure CPU version.

\paragraph{CPU}
On the CPU we perform the inner product by element-wise polynomial-polynomial multiplications using the implementation of the polynomial.
This operation can easily parallelized using OpenMP by splitting the vector and distributing equal-sized chunks among the available threads.
Since all polynomials are of the same structure the work is well balanced between the threads.

The highly parallel structure of the operation suggests a SIMD approach.
However to perform 4 large integer operations in parallel on current \verb|x86-64| architectures is not advisable,
since the kernels need to be based 32 bit integer operations instead of 64 bit operations which are available for the sequential version.
This increases the number of required multiply instructions for a large integer multiplication by a factor of 4.
Using current Streaming SIMD Extensions (SSE) allows only two 32bit multiplications yielding 64bit results in a SIMD fashion.
Thus a SIMD implementation using SSE will be two times slower than our serial version using 64 bit integers.
Furthermore we neglected the carry bit propagation, which needs to be done manually in SSE,
while the sequential version takes advantage of the hardware support for carry bit propagation.

The upcoming AVX2 instruction set will feature integer SIMD operations with twice the vector width of SSE. However, it will neither support any 64 bit integer operations nor any carry bit propagation. Therefore it will at most be as fast as our sequential version.

%\begin{itemize}
%\item We use the single large integer kernels
%\item OpenMP parallelization: split vectors into chunks, each thread does part of the inner product. Very regular problem, good load balance.
%\item A SIMD implementation to perform 4 large integer operations in parallel on current \verb|x86_64| architectures is not useful, since we had to use 32bit integer operations, which increases the number of multiply instructions for a large integer multiplication by a factor of 4 compared to the 64bit based implementation.
%Using current Streaming SIMD Extensions (SSE) allows only two 32bit multiplications yielding 64bit results in a SIMD fashion.
%Thus a SIMD implementation using SSE will be two times slower than our serial version using 64 bit integers.
%We also neglected the carry bit propagation, which needs to be done manually in SSE, while the sequential version takes advantage of the hardware support for carry bit propagation.
%\item The upcoming AVX2 will double the vector width and support four 32bit integer multiplications, but no carry bit support.
%\end{itemize}
\paragraph{GPU}
\begin{itemize}
\item Many small tasks of the same form. $\rightarrow$ well suited for GPUs.
\item {\bf Implementation details:}
\item based on 32bit integers due to architecture.
\item Element-wise polynomial multiplication + reduction at the end.
\item Each thread calculates one coefficient $c_{IJKL}$ the product polynomial.
\item Several coefficients $a_{ijkl}$, $b_{i'j'k'l'}$ of the polynomials to be multiplied contribute to the same coefficient $c_{IJKL}$
\begin{equation}
    c_{IJKL} = \sum_{i,j,k,l} a_{ijkl} \cdot b_{I-i,J-j,K-k,L-l}
\end{equation}
\item The number of terms in this sum depends on the orders $I,J,K,L$ of the resulting coefficient $\rightarrow$ load in-balance
%\item {\color{blue} We design an execution plan. We group the coefficients of a fictive result polynomial by packet of 32 } into a task and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}).
\item Since threads within a warp are executed in a lockstep manner, we try to give all threads within a warp the same amount of work. This way the threads have to wait as little as possible for the other threads of the warp to finish.
\item We group the coefficients by 32 into a ``task'' and distribute them among the warps of threads (Fig.\ \ref{fig:GPU_load_balance}), i.e.\ each thread calculates one coefficient of the resulting polynomial.
\begin{figure}[t]
    \begin{center}
    \mbox{
        \includegraphics[scale=0.37, angle=-90]{coeffs.eps} 
        \includegraphics[scale=0.37, angle=-90]{warp.eps} 
    }
    \caption{Left: Number of contributions to each coefficient of the resulting polynomial (red line) sorted in reverse order and split into tasks. Right: Load balancing of the tasks over the warps.}
    \label{fig:GPU_load_balance}
    \end{center}
\end{figure}
\item Each thread of the warp gets the indices of the resulting coefficient and determines the necessary terms to calculate the output coefficient.
\item Reduction is implemented according to the best practices by NVIDIA (TODO reference)
%\item {\color{blue} This algorithm is efficient whatever the type of polynomial. But the termination of the contributions of an output coefficient can be tricky}
%\item Coalesced memory access?
\item We use the Karatsuba algorithm \cite{Karatsuba1963} for the largest extended multiplication (256 bit to 512 bit) on GPU, because an integer addition requires substantially less cycles than multiplication the GPU.
\item Using NVIDIA parallel thread execution (PTX) assembly language to benefit from hardware carry bit propagation \verb|addc.cc| and multiply-add with carry bit propagation \verb|madc.cc|.
\end{itemize}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\section{Conclusion}
\begin{itemize}
\item TODO
\item Extend integers up to 2048 bits.
\item Allow more polynomial structures
\item License?
\end{itemize}
